Characterizing edges in signed and vector-valued graphs

### 2017-12-16 intro/learning.tex
pass over the intro

  \item[community detection~\autocite{FortunatoSurvey10}] The goal is to cluster nodes in tightly
    connected groups that are loosely connected with the rest of the graph. This allows for a better
    understanding of the graph organization, and present a higher level view than looking at
    individual nodes relationships. For instance, it has been used to identify proteins functional
    groups~\autocite{clusterBio03} or see how different scientific fields relate based on
    publication data~\autocite{clusterScience08}.

@@

  \item[network evolution~\autocite{networkEvolution14}] These methods focus on the mechanisms and
    consequences of the growth of networks, as well as how to keep the results of some data mining
    algorithms up to date and relevant. Monitoring the changes in the interactions of proteins
    can indeed be used as an early indicator for some kind of diabetes~\autocite{evolBio10}.
    Furthermore, sudden changes in a network of computers are usually worth investigating, for they
    might signal external attacks~\autocite{evolSecurity04}.

>>>>

  \item[community detection~\autocite{FortunatoSurvey10}] The goal is to cluster nodes in tightly
    connected groups that are loosely connected with the rest of the graph. This allows for a better
    understanding of the graph organization, and present a higher level view than looking at the
    individual node relationships. For instance, it has been used to identify proteins functional
    groups~\autocite{clusterBio03}, or to see how different scientific fields relate based on
    publication data~\autocite{clusterScience08}.

@@

  \item[network evolution~\autocite{networkEvolution14}] These methods focus on the mechanisms and
    consequences of the growth of networks. They also seek ways to keep the results of some data mining
    algorithms up to date and relevant. Monitoring the changes in the interactions of proteins
    can indeed be used as an early indicator for some kind of diabetes~\autocite{evolBio10}.
    Furthermore, sudden changes in a network of computers are usually worth investigating, for they
    might signal external attacks~\autocite{evolSecurity04}.

### 2017-12-16 intro/main.tex
pass over the intro

Graphs are a natural way to represent the relationships over a set of entities. Because of the
simplicity and flexibility of this formalism, graphs are ubiquitous and have been used in countless
fields. To make the rest of our explanations more concrete, we now give three examples of graphs
that we consider later on.
\begin{enumerate}[nosep]%,leftmargin=*]
\item In a social network\footnote{In the rest of this thesis, we use the terms \emph{graph} and
\emph{network} interchangeably.}, the nodes of the graph are human users, and relationships
between nodes denote interactions between the corresponding users. One social network we study
in \autoref{chap:troll} summarizes the connections between the editors of \wik{}. These editors
can be promoted to administrators after a vote from their peers. An edge \euv{} in this network
means that user $u$ has voted on the possible promotion of $v$.
\item In computer vision, we can represent an image as a graph. Each pixel is a node, and those
nodes are connected to four neighbors, namely the adjacent pixel from the top, bottom, left and
right sides. The relationship in this graph is therefore adjacency in an image.
\item In e-commerce, we can consider a co-purchase network. Nodes are products being sold on a
website, and two nodes are connected if the corresponding products are frequently bought
together by customers, for instance a phone and a memory card.
\end{enumerate}

Not only are there graphs in many domains, but the progress of technology in the last few decades
has made it easier to collect many graphs in every single domain, sometimes with up to billion of
nodes and hundred of billions of edges. The availability of such large amount of structured data has
prompted the development of automated methods to extract insights from them. For instance, it is
possible to cluster nodes into coherent groups, predict the category of the nodes or study how to
best propagate information within a graph. We describe in more details these possibilities and
others in \autoref{sec:learning_in_graphs}. By doing so, our goal is to illustrate the wide
potential of \emph{learning in graphs}.

@@

At this point of the discussion though, we have only considered the most common and simple kind of
graph, one representing a single type of relation between nodes.\footnote{Note that by \enquote{type
of relation}, we do not refer to some graph-theoretical characteristic of an edge such as being directed or weighted.
Indeed, all the graphs we consider have homogeneous edges with that respect. Instead we mean domain
specific semantic, as we will make clear in examples.} However, in many situations, there
are two dimensions along which graphs exhibit more complexity. First, nodes have more than one type
of relation among each other. Second, two nodes are not only connected because of their global
similarity but also for more nuanced reasons. As a case in point, let us look again at our
three examples. In the \wik{} network, a vote can support the promotion or oppose it. This
additional information enrich our understanding of the relationships between the nodes in the graph.
It also points out that two editors can be connected because they share a common topic of interest,
but come from different socio-economic backgrounds and therefore cannot agree on this topic.
Likewise in an image, an edge between two pixels is positive if the two pixels belong to the same
object (say a car or a building) and negative otherwise. This information could be used to segment
the image. In the co-purchase network, there are even more than two types of relation. Assuming the
products are movies, two movies can be frequently bought together because they are part of a series
(like Star Wars), because they have the same director but different genre, because they receive the
same prestigious awards in different years, and so on. The second type of relation (\enquote{same
director} and \enquote{different genre}) is actually an example of a mixed relation, for it mixes
similarity and dissimilarity.

@@

It is natural to expect that we can extract finer insights from graphs with edge semantics. However,
in many cases, the information about edge types is not available, at least not in a convenient form.
We therefore present in
\autoref{sec:predicting_edge_type} the problem of characterizing edges in complex networks.
Informally, given an input graph and possibly some extra information, we want to predict the type of
every edge. This problem can take several forms depending of what information is available as input.
First, the graph may be directed or undirected. Second, there might be two edge types (in the case of
signed graphs) or more (for general multilayer graphs). Third, the problem can be supervised or not.
In the supervised setting, we are provided with labels for some of the edges. In the unsupervised
setting, there is no label at all. We thus make the additional assumption that we observe some
\emph{attributes} of the nodes. Once again, we return to our three examples to illustrate what
those attributes can be. In the \wik{} graph, we could have for each user data about age, experience and
area of expertise. In an image, each pixel is associated with a color, along with higher order
visual features. Each product of a co-purchase network comes with information about, \eg{}, its price,
popularity, category and availability. In the absence of label, our intuition is that these
attributes can inform us about the type of edge between nodes.

>>>>

Graphs are a natural way to represent the relationships over a set of entities. Because of the
simplicity and flexibility of this formalism, graphs are ubiquitous and have been used in countless
fields. To make the rest of our explanations more concrete, we now give three examples of graphs
that we consider later on.
\begin{enumerate}[nosep]%,leftmargin=*]
\item In a social network\footnote{In the rest of this thesis, we use the terms \emph{graph} and
\emph{network} interchangeably.}, the nodes of the graph are human users, and relationships
between nodes denote interactions between the corresponding users. One social network we study
in \autoref{chap:troll} summarizes the connections between the editors of \wik{}. These editors
can be promoted to administrators after a vote from their peers. An edge \euv{} in this network
means that user $u$ has voted on the possible promotion of $v$.
\item In computer vision, we can represent an image as a graph. Each pixel is a node, and those
nodes are connected to four neighbors, namely the adjacent pixels from the top, bottom, left and
right sides. The relationship in this graph is therefore adjacency in an image.
\item In e-commerce, we can consider a co-purchase network. Nodes are products being sold on a
website, and two nodes are connected if the corresponding products are frequently bought
together by customers, for instance a phone and a memory card.
\end{enumerate}

Not only are there graphs in many domains, but the progress of technology in the last few decades
has made it easier to collect many graphs in every single domain, sometimes with up to billions of
nodes and hundreds of billions of edges. The availability of such large amounts of structured data has
prompted the development of automated methods to extract insights from them. For instance, it is
possible to cluster nodes into coherent groups, predict the category of the nodes or study how to
best propagate information within a graph. We describe in more details these possibilities and
others in \autoref{sec:learning_in_graphs}. By doing so, our goal is to illustrate the wide
potential of \emph{learning in graphs}.

@@

At this point of the discussion though, we have only considered the most common and simple kind of
graph, one representing a single type of relation between nodes.\footnote{Note that by \enquote{type
of relation}, we do not refer to some graph-theoretical characteristic of an edge, such as being directed or weighted.
Indeed, all the graphs we consider have homogeneous edges with that respect. Instead, we mean \emph{domain
specific semantic}, as we will make clear in examples.} However, in many situations, there
are two dimensions along which graphs exhibit more complexity. First, nodes have more than one type
of relation among each other. Second, two nodes are not only connected because of their global
similarity but also for more nuanced reasons. As a case in point, let us look again at our
three examples. In the \wik{} network, a vote can support the promotion or oppose it. This
additional information enrich our understanding of the relationships among the nodes in the graph.
It also points out that two editors can be connected because they share a common topic of interest,
but come from different socio-economic backgrounds and therefore cannot agree on this topic.
Likewise in an image, an edge between two pixels is positive if the two pixels belong to the same
object (say a car or a building) and negative otherwise. This information could be used to segment
the image. In the co-purchase network, there are even more than two types of relation. Assuming the
products are movies, two movies can be frequently bought together because they are part of a series
(like Star Wars), because they have the same director but different genre, because they receive the
same prestigious award in different years, and so on. The second type of relation (\enquote{same
director} and \enquote{different genre}) is actually an example of a mixed relation, for it combines
similarity and dissimilarity over several features of the nodes.

@@

It is natural to expect that we can extract finer insights from graphs with edge semantics. However,
in many cases, the information about edge types is not available, at least not in a convenient form.
We therefore present in
\autoref{sec:predicting_edge_type} the problem of characterizing edges in complex networks.
Informally, given an input graph and possibly some extra information, we want to predict the type of
every edge. This problem can take several forms depending on what information is available as input.
First, the graph may be directed or undirected. Second, there might be two edge types (in the case of
signed graphs) or more (for general multilayer graphs). Third, the problem can be supervised or not.
In the supervised setting, we are provided with labels for some of the edges. In the unsupervised
setting, there is no label at all. We thus make the additional assumption that we observe some
\emph{attributes} of the nodes. Once again, we return to our three examples to illustrate what
those attributes can be. In the \wik{} graph, we could have for each user data about her age, experience and
area of expertise. In an image, each pixel is associated with a color, along with higher order
visual features. Each product of a co-purchase network comes with information about, \eg{}, its price,
popularity, category and availability. In the absence of label, our intuition is that these
attributes can inform us about the types of edge among nodes.

### 2017-12-16 intro/outline.tex
pass over the intro

  \item We start in \autoref{chap:troll} by addressing the \esp{} problem in \dssn{}. Our goal is to
    design a method that is scalable, principled and accurate. For that, we start by introducing a
    generative model for signs, and derive approximations of the optimal Bayes predictor or the
    maximum likelihood predictor. We confirm the theoretical soundness of these approaches by
    performing extensive synthetic and real world experiments. Finally, to the best of our
    knowledge, we are the first to give an online algorithm for the \esp{} problem on directed
    graphs. This chapter is based on an existing publication~\autocite{trollSign17}.

>>>>

  \item We start in \autoref{chap:troll} by addressing the \esp{} problem in \dssn{}. Our goal is to
    design a method that is scalable, principled and accurate. For that, we start by introducing a
    generative model for signs, and derive approximations of the optimal Bayes predictor and of the
    maximum likelihood predictor. We confirm the theoretical soundness of these approaches by
    performing extensive synthetic and real world experiments. Finally, to the best of our
    knowledge, we are the first to give an online algorithm for the \esp{} problem on directed
    graphs. This chapter is based on an existing publication~\autocite{trollSign17}.

### 2017-12-16 intro/predicting.tex
pass over the intro

  \item There exists a function that can perfectly label any edge because it is tailored to this
    specific graph, but calling it is expensive. For instance, say we want to know whether two
    connected users of a social network are tied to each other through family, work, school or
    hobby. The labeling function in this case simply asks users to label their relationships,
    assuming the answers do not contain any noise. At the scale of Facebook, this would require
    asking hundred of questions to each user on average, which is quite time consuming. Furthermore,
    it would also cost marketing resource to convince users this is beneficial for them, and
    engineering time to ensure this information remain confidential. Likewise, in a biological
    network, determining whether two proteins interact positively or negatively with each other is
    achieved by a lab experiment, which requires time and material.

@@

  \item The same \esp{} problem, where the input graph is undirected. In this case, we consider an
    active variant where instead of being given a random training set of labels, we have a budget of
    queries we can use to request arbitrary labels.

@@

We now review existing approaches addressing these problems, in order to highlight our contribution
in the next section.

The modern formulation of the \esp{} on directed graphs can be attributed to
\textcite{Leskovec2010}. Their idea is to compute local features of the nodes based on the training
labels. Such features include variations on the node degree, such as the number of positive
outgoing edges or the number of negative incoming edges. These node features are combined into edge
features, and a supervised classification algorithm is trained. Several works have devised
additional features~\autocites{Bayesian15}{Yuan2017}, most notably based on scoring the nodes in a
Page Rank fashion~\autocites{traag2010exponential}{shahriari2014ranking}{wu2016troll}. These methods
are accurate, fast ---for they are local, and interpretable ---for the features are hand crafted.
The drawback is that this feature engineering is done mostly in an ad hoc way. On the other hand, it
is also possible to look at the problem from a global point of view, by completing the
signed adjacency matrix through low rank
factorization~\autocites{LowRankCompletion14}{OnlineCompletion17}. These methods are also accurate
and bypass feature engineering, but by nature, they require careful algorithms to scale with larger
networks. Finally, during the writing of this manuscript, several papers have been published, which
use embedding of the nodes in a low dimensional space, based on the training
labels~\autocites{SIGNet17}{SNE17}. This also avoids feature engineering but is global in nature and
not so interpretable.

One way to solve the problem on undirected graphs would be to apply the previous methods, having
first replaced every edge $(u,v)$ by two directed edges \euv{} and \evu{}. This is not quite
satisfactory, both from a theoretical but also practical point of view (see
\autoref{sub:need_for_a_directed_graph}). Instead, \textcite{Cesa-Bianchi2012b} draw a connection
between the \esp{} problem and the \pcc{} problem. Recall that given a fully labeled signed graph,
the solution to \pcc{} is a partition of the nodes that minimize the number of disagreement
edges~\autocite{Bansal2002}. \Textcite{Cesa-Bianchi2012b} assume that the signs are originally
consistent with an underlying, hidden $2$-clustering, but we only observe the signs after they have
been flipped \uar{}. In this case, they show that the optimal number of disagreements is a lower
bound of the number of mistakes of any active \esp{} algorithm. One natural approach would then be
to solve \pcc{} based on the observed signs, and predict the remaining signs consistently with the
inferred clusters. At first it seems hopeless, as \pcc{} is difficult to approximate on general
graphs~\autocite{Charikar2003}, even when there are only two clusters~\autocite{Giotis2006}.
However, this worst case analysis does not forbid more positive results on signed graphs that
exhibit stability under perturbation~\autocites{clusteringFeasibility15}{StableCC09}{StableLP09} or
are obtained through perturbations from an ideal case~\autocites{plantedAilon09}{Makarychev2014}.
Furthermore, in the active setting, the learner gets to choose which signs are observed. The general
idea of \textcite{Cesa-Bianchi2012b} is thus to use the query budget to build fully labeled paths
between each connected nodes. Those paths must be as short as possible, since the predicted sign of
$(u,v)$ is the product of the signs along the path from $u$ to $v$. Therefore, the shorter the path
and the less influence of the random perturbations. At the same times, those paths must span the
whole graph. This is the topic of an active research area~\autocites{Abraham2012}{Spanner17}.

As for the \ecp{} problem, to the best of our knowledge, it has not been studied under our
assumptions. A more common problem in attributed graphs is to cluster nodes into
communities~\autocites{LeskovecEgo12}{Yang2013}{Xu2014}{ZhangModelFree16}. However, it is not
immediate how such methods, generally based on generative models, could be adapted to our problem.
Direct approaches to classify edges have been proposed, based on graphical
models~\autocite{graphicalModelTies11}, nearest neighbors with a customized
distance~\autocite{Aggarwal2016a} and edge embedding in knowledge graphs~\autocite{transE13}. Yet
they all rely on having training labels as supervision. A simple unsupervised method is to cluster
the line graph of the input graph~\autocite{LineGraph09}, but this does not take advantage of
attributes. Another method uses topological features~\autocite{ahmed2017roles} but is rather complex
and thus not very interpretable. If we further assume that each edge type is associated with an
Euclidean space, and that the position of nodes in one space are not correlated with their position
in another space, then it is possible to recover an approximation of these Euclidean metrics in
polynomial time without any supervision~\autocite{Abraham2012a}.

>>>>

  \item There exists a function that can perfectly label any edge because it is tailored to this
    specific graph, but calling it is expensive. For instance, say we want to know whether two
    connected users of a social network are tied to each other through family, work, school or
    hobby. The labeling function in this case simply asks users to label their relationships,
    assuming the answers do not contain any noise. At the scale of Facebook, this would require
    asking hundred of questions to each user on average, which is quite time-consuming. Furthermore,
    it would also cost marketing resource to convince users this is beneficial for them, and
    engineering time to ensure this information remain confidential. Likewise, in a biological
    network, determining whether two proteins interact positively or negatively with each other is
    achieved by a lab experiment, which requires time and material.

@@

  \item The same \esp{} problem, where the input graph is undirected. In this case, we consider an
    active variant where, instead of being given a random training set of labels, we have a budget of
    queries we can use to request arbitrary labels.

@@

We now briefly review existing approaches addressing these problems, in order to highlight our contribution
in the next section.

The modern formulation of the \esp{} on directed graphs can be attributed to
\textcite{Leskovec2010}. Their idea is to compute local features of the nodes based on the training
labels. Such features include variations on the node degree, such as the number of positive
outgoing edges or the number of negative incoming edges. These node features are combined into edge
features, and a supervised classification algorithm is trained. Several works have devised
additional features~\autocites{Bayesian15}{Yuan2017}, most notably based on scoring the nodes in a
Page Rank fashion~\autocites{traag2010exponential}{shahriari2014ranking}{wu2016troll}. These methods
are accurate, fast ---for they are local, and interpretable ---for the features are hand crafted.
The drawback is that this feature engineering is done mostly in an ad hoc way. On the other hand, it
is also possible to look at the problem from a global point of view, by completing the
signed adjacency matrix through low rank
factorization~\autocites{LowRankCompletion14}{OnlineCompletion17}. These methods are also accurate,
and bypass feature engineering, but by nature, they require careful algorithms to scale with larger
networks. Finally, during the writing of this manuscript, several papers have been published, which
use embedding of the nodes in a low dimensional space, based on the training
labels~\autocites{SIGNet17}{SNE17}. This also avoids feature engineering but is global in nature and
not so interpretable.

One way to solve the problem on undirected graphs would be to apply the previous methods, having
first replaced every edge $(u,v)$ by two directed edges \euv{} and \evu{}. This is not quite
satisfactory, both from a theoretical but also practical point of view (see
\autoref{sub:need_for_a_directed_graph}). Instead, \textcite{Cesa-Bianchi2012b} draw a connection
between the \esp{} problem and the \pcc{} problem. Recall that given a fully labeled signed graph,
the solution to \pcc{} is a partition of the nodes that minimize the number of disagreement
edges~\autocite{Bansal2002}. \Textcite{Cesa-Bianchi2012b} assume that the signs are originally
consistent with an underlying, hidden $2$-clustering, but that we only observe the signs after they have
been flipped \uar{}. In this case, they show that the optimal number of disagreements is a lower
bound of the number of mistakes made by any active \esp{} algorithm. One natural approach would then be
to solve \pcc{} based on the observed signs, and predict the remaining signs consistently with the
inferred clusters. At first, it seems hopeless, as \pcc{} is difficult to approximate on general
graphs~\autocite{Charikar2003}, even when there are only two clusters~\autocite{Giotis2006}.
However, this worst case analysis does not forbid more positive results on signed graphs that
exhibit stability under perturbation~\autocites{clusteringFeasibility15}{StableCC09}{StableLP09} or
are obtained through perturbations from an ideal case~\autocites{plantedAilon09}{Makarychev2014}.
Furthermore, in the active setting, the learner gets to choose which signs are observed. The general
idea of \textcite{Cesa-Bianchi2012b} is thus to use the query budget to build fully labeled paths
between each connected nodes. Those paths must be as short as possible, since the predicted sign of
$(u,v)$ is the product of the signs along the path from $u$ to $v$. Therefore, the shorter the path
and the less influence of the random perturbations. At the same time, those paths must span the
whole graph. This is the topic of an active research area~\autocites{Abraham2012}{Spanner17}.

As for the \ecp{} problem, to the best of our knowledge, it has not been studied under our
assumptions. A more common problem in attributed graphs is to cluster nodes into
communities~\autocites{LeskovecEgo12}{Yang2013}{Xu2014}{ZhangModelFree16}. However, it is not
immediate how such methods, generally based on generative models, could be adapted to our problem.
Direct approaches to classify edges have been proposed, based on graphical
models~\autocite{graphicalModelTies11}, nearest neighbors with a customized
distance~\autocite{Aggarwal2016a} and edge embedding in knowledge graphs~\autocite{transE13}. Yet
they all rely on having training labels as supervision. A simple unsupervised method is to cluster
the line graph of the input graph~\autocite{LineGraph09}, but this does not take advantage of
attributes. Another method uses topological features~\autocite{ahmed2017roles} but is rather complex
and thus not very interpretable. If we further assume that each edge type is associated with a
Euclidean space, and that the position of nodes in one space are not correlated with their position
in another space, then it is possible to recover an approximation of these Euclidean metrics in
polynomial time without any supervision~\autocite{Abraham2012a}.

### 2017-12-16 intro/semantics.tex
pass over the intro

The three graphs we described in the introduction are examples of what we call complex networks.
Such graphs have more than one type of edge, and edges are not explained by mere global node
similarity. The point of this section is to illustrate that this notion is not simply a mathematical
variation of a \enquote{simple} graph, where the adjacency matrix would take values in $\{-1, 0, 1\}$ or
$\{0, 1, \ldots, k\}$ instead of $\{0, 1\}$. Rather, we reviews many applications in various domains
where complex networks are the right model to represent rich networked data. First, we showcase many
uses of signed networks. In that case, we do not insist on the fact that similarity is not the only
driver of node connections, for it is implicit that negative edges actually denote dissimilarity.
Second, we present multilayer graphs, that is graphs with more than two types of edges. When
applicable, we highlight how the connections in such graphs indeed rely on a nuanced similarity or
dissimilarity across some of the nodes attributes.

To elaborate on this last point, our hypothesis is that connections in complex networks are the
results of both partial homophily (that is, nodes are connected when they are similar on a subset of
the attributes) and partial heterophily (that is, nodes are connected when they are dissimilar on a
subset of the attributes). As examples of the latter, think of dating websites ---where most users
are linked with users of the opposite gender~\autocites{homophilyMyspace09}{Tinder16}; diffusing
innovations ---where meeting people with different backgrounds and point of views is crucial to
favor diversity and creativity~\autocite{rogers2003diffusion}; and online news consumption ---where
connecting people from different sides of the political spectrum helps to avoid echo chambers and
fuel a democratic debate~\autocite{balancedNews17}.

@@

Besides signed graphs, in this thesis we also consider at multilayer
graphs~\autocites{Kivela2014}{multiSurvey14}. Those are graphs with $k$ edge types, and the name
refer to the fact we can see them as the superposition of the $k$ subgraphs induced by each edge
type. Even when $k=2$, we make a distinction between signed graphs and multilayer graphs. In signed
graphs, the two types of edge have reverse semantic, whereas in a 2-layers graph, they simply
denotes two possible interactions, for instance advisor-advisee or regular coauthors relation in a
citation network~\autocite{Advisor10}. In general though, we focus on cases where $k$ is larger than
$2$, and not larger than a few dozen in order to preserve interpretability.
Like signed graphs, these multilayer graphs are versatile enough to be used in many fields.

\paragraph{Social networks} \Textcite{Szell2010} model the interactions of the players of
\emph{Pardus}, a massively multilayer online game. These players can be friend or enemy, send
private messages, trade resources, attack each other and set a bounty on the head of another
player. These six types of interactions are either positive or negative, but their nuances cannot
simply by explained by global similarity and dissimilarity. Another examples are the users of the
photo sharing website Flickr.  They can interact in eleven ways, either directly or through
comments, shared tags, groups membership and so on~\autocite{RecoFlickrMulti11}. Again, while being
part of a common group denote shared interests between two users, overall they must also differ in
some other attributes (for instance location) in order to bring diversity to this group.

@@

\paragraph{Economic networks} In our ever increasingly globalized economy, entities around the world
are getting more and more tightly connected. However, those connections take on many different
forms simultaneously. For something seemingly as simple as the connections
between the largest \np{951} ports in the world, one must already notice that these connections can be
implemented by any combinations of three kinds of ships: bulk dry carriers, container ships and oil
tankers~\autocite{ports3kindofships10}. Likewise, the 162 countries of the International Trade
Network are connected by 96 kinds of commodities they can exchange~\autocite{worldTradeNetwork10}.
Finally, \textcite{KantPeace15} studies international relations through the lens of Kant three
folded program for peace,  based on democracy, economic dependence and supra national governance.
They build the graph of all countries and connect them in three layers. All democracies form a
clique in the first layer, countries are connected with weight proportional to the amount of yearly
trade in the second layer and with weight proportional to the number of international organisations
they belong to together in the third layer. While the essence of commerce is to exploit differential
between the partners involved (\ie{} heterophily), trade (and in the last example participating in
common institution) intensity also involve geographical, historical and cultural ties (\ie{}
homophily).

\paragraph{Biological networks} The interactions in biology also take several forms and studying
them as a whole has proved fruitful. One example is a genes co-expression network, where each
connection was tested under 130 different experimental conditions, providing as many
layers~\autocite{bioLayerExp11}.
Multilayer graphs are also a relevant way to represent ecological
networks~\autocite{EcologyMultiReview17}. For instance, \textcite{EcoChile15} build the graph of
more than 100 species living on the Chilean
coast\footnote{\url{http://app.mappr.io/play/chile-marine-intertidal-network}} and divide their
interactions in three categories: trophic (that is one specie eats another one), positive
non-trophic (\eg{} refuge providing) and negative non-trophic (\eg{} competition for shared
resources).
Finally, in neuroscience, multilayer graphs have recently emerged as a useful tool to better
understand the human brain~\autocite{Neuroscience16}. The nodes of such graphs are neurons, and the
edges can be labeled in various ways: some correspond to actual physical links while others are
functional ---\ie{} neurons responding in the same way to external stimuli, some are present in
healthy subject and others in patient cohort, some are acquired through MRI and others by EEG.

>>>>

The three graphs we described in the introduction are examples of what we call complex networks.
Such graphs have more than one type of edge, and edges are not explained by mere global node
similarity. The point of this section is to illustrate that this notion is not simply a mathematical
variation of a \enquote{simple} graph, where the adjacency matrix would take values in $\{-1, 0, 1\}$ or
$\{0, 1, \ldots, k\}$ instead of $\{0, 1\}$. Rather, we review many applications in various domains
where complex networks are the right model to represent rich networked data. First, we showcase many
uses of signed networks. In that case, we do not insist on the fact that similarity is not the only
driver of node connections, for it is implicit that negative edges actually denote dissimilarity.
Second, we present multilayer graphs, that is graphs with more than two types of edges. When
applicable, we highlight how the connections in such graphs indeed rely on a nuanced similarity or
dissimilarity across some of the nodes attributes.

To elaborate on this last point, our hypothesis is that connections in complex networks are the
results of both partial homophily (that is, nodes are connected when they are similar on a subset of
the attributes) and partial heterophily (that is, nodes are connected when they are dissimilar on a
subset of the attributes). As examples of the latter, think of dating websites ---where most users
are linked with users of the opposite gender~\autocites{homophilyMyspace09}{TinderDesc16}; diffusing
innovations ---where meeting people with different backgrounds and point of views is crucial to
favor diversity and creativity~\autocite{rogers2003diffusion}; and online news consumption ---where
connecting people from different sides of the political spectrum helps to avoid echo chambers and
instead fuel a democratic debate~\autocite{balancedNews17}.

@@

Besides signed graphs, in this thesis we also consider multilayer
graphs~\autocites{Kivela2014}{multiSurvey14}. Those are graphs with $k$ edge types, and the name
refer to the fact we can see them as the superposition of the $k$ subgraphs induced by each edge
type. Even when $k=2$, we make a distinction between signed graphs and multilayer graphs. In signed
graphs, the two types of edge have reverse semantic, whereas in a 2-layers graph, they simply
denote two possible interactions, for instance advisor-advisee or regular coauthors relation in a
citation network~\autocite{Advisor10}. In general though, we focus on cases where $k$ is larger than
$2$, and not larger than a few dozen in order to preserve interpretability.
Like signed graphs, these multilayer graphs are versatile enough to be used in many fields.

\paragraph{Social networks} \Textcite{Szell2010} model the interactions of the players of
\emph{Pardus}, a massively multilayer online game. These players can be friend or enemy, send
private messages, trade resources, attack each other and set a bounty on the head of another
player. These six types of interactions are either positive or negative, but their nuances cannot
simply by explained by global similarity and dissimilarity. Another example is photo sharing website
Flickr and its users. They can interact in eleven ways, either directly or through
comments, shared tags, groups membership and so on~\autocite{RecoFlickrMulti11}. Again, while being
part of a common group denotes shared interests between two users, overall they must also differ in
some other attributes (for instance location) in order to bring diversity to this group.

@@

\paragraph{Economic networks} In our ever increasingly globalized economy, entities around the world
are getting more and more tightly connected. However, those connections take on many different
forms simultaneously. For something seemingly as simple as the connections
between the largest \np{951} ports in the world, one must already notice that these connections can be
implemented by any combinations of three kinds of ships: bulk dry carriers, container ships and oil
tankers~\autocite{ports3kindofships10}. Likewise, the 162 countries of the International Trade
Network are connected by 96 kinds of commodities they can exchange~\autocite{worldTradeNetwork10}.
Finally, \textcite{KantPeace15} studies international relations through the lens of Kant's three
folded program for peace,  based on democracy, economic dependence and supra national governance.
They build the graph of all countries and connect them in three layers. All democracies form a
clique in the first layer, countries are connected with weight proportional to the amount of yearly
trade in the second layer and with weight proportional to the number of international organizations
they belong to together in the third layer. While the essence of commerce is to exploit differential
between the partners involved (\ie{} heterophily), trade intensity and participating in
common institutions also involve geographical, historical and cultural ties (\ie{}
homophily).

\paragraph{Biological networks} The interactions in biology also take several forms and studying
them as a whole has proved fruitful. One example is a genes co-expression network, where each
connection was tested under 130 different experimental conditions, providing as many
layers~\autocite{bioLayerExp11}.
Multilayer graphs are also a relevant way to represent ecological
networks~\autocite{EcologyMultiReview17}. For instance, \textcite{EcoChile15} build the graph of
more than 100 species living on the Chilean
coast\footnote{\url{http://app.mappr.io/play/chile-marine-intertidal-network}} and divide their
interactions in three categories: trophic (\ie{} one specie eating another one), positive
non-trophic (\eg{} refuge providing) and negative non-trophic (\eg{} competition for shared
resources).
Finally, in neuroscience, multilayer graphs have recently emerged as a useful tool to better
understand the human brain~\autocite{Neuroscience16}. The nodes of such graphs are neurons, and the
edges can be labeled in various ways: some correspond to actual physical links while others are
functional (\ie{} neurons responding in the same way to external stimuli), some are present in
healthy subjects and others in treated patients, some are acquired through MRI and others by EEG.
